Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories

### 2020-10-09 paper/main.tex
removed duplicate page numbering in main text and supplement

\begin{document}
\begin{titlepage}
  \maketitle
  \end{titlepage}

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/searchlights}
\caption{\small \textbf{Brain structures that underlie the transformation of experience into memory.} \textbf{A.} We isolated the proximal diagonals from the upper triangle of the episode correlation matrix, and applied this same diagonal mask to the voxel response correlation matrix for each cube of voxels in the brain. We then searched for brain regions whose activation timeseries consistently exhibited a similar proximal correlational structure to the episode model, across participants.  \textbf{B.} We used dynamic time warping \citep{BernClif94} to align each participant's recall timeseries to the TR timeseries of the episode.  We then applied the same diagonal mask used in Panel A to isolate the proximal temporal correlations and searched for brain regions whose activation timeseries for an individual consistently exhibited a similar proximal correlational structure to each individual's recall.  \textbf{C.} We identified a network of regions sensitive to the narrative structure of participants' ongoing experience.  The map shown is thresholded at $p < 0.05$, corrected.  The top ten \texttt{Neurosynth} terms displayed in the panel were computed using the unthresholded map.  \textbf{D}. We also identified a network of regions sensitive to how individuals would later structure the episode's content in their recalls.  The map shown is thresholded at $p < 0.05$, corrected.  The top ten \texttt{Neurosynth} terms displayed in the panel were computed using the unthresholded map.}
\label{fig:brainz}
\end{figure}

@@

The neural analyses we carried out (Fig.~\ref{fig:brainz}) also leveraged our geometric framework for characterizing the shapes of the episode and participants' recountings.  We identified one network of regions whose responses tracked with temporal correlations in the conceptual content of the episode (as quantified by topic models applied to a set of annotations about the episode).  This network included orbitofrontal cortex, ventromedial prefrontal cortex, and striatum, among others.  As reviewed by \cite{RangRitc12}, several of these regions are members of the \textit{anterior temporal system}, which has been implicated in assessing and processing the familiarity of ongoing experiences, emotions, social cognition, and reward.  A second network we identified tracked with temporal correlations in the idiosyncratic conceptual content of participants' subsequent recountings of the episode.  This network included occipital cortex, extrastriate cortex, fusiform gyrus, and the precuneus.  Several of these regions are members of the \textit{posterior medial system}~\citep{RangRitc12}, which has been implicated in matching incoming cues about the current situation to internally maintained \textit{situation models} that specify the parameters and expectations inherent to the current situation~\citep[also see][]{ZackEtal07, ZwaaRadv98}.  Taken together, our results support the notion that these two (partially overlapping) networks work in coordination to make sense of our ongoing experiences, distort them in a way that links them with our prior knowledge and experiences, and encodes those distorted representations into memory for our later use.

Our general approach draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  Our approach to identifying neural responses to naturalistic stimuli (including experiences) entails building an explicit model of the stimulus dynamics and searching for brain regions whose responses are consistent with the model~\citep[also see][]{HuthEtal12, HuthEtal16}.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have presented a clever alternative approach: rather than building an explicit stimulus model, these studies instead search for brain responses to the stimulus that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time~\citep[also see][]{SimoChan20}.  These purely brain-driven approaches are well suited to identifying which brain structures exhibit similar stimulus-driven responses across individuals.  Further, because neural response dynamics are observed data (rather than model approximations), such approaches do not require a detailed understanding of which stimulus properties or features might be driving the observed responses.  However, this also means that the specific stimulus features driving those responses are typically opaque to the researcher.  Our approach is complementary.  By explicitly modeling the stimulus dynamics, we are able to relate specific stimulus features to behavioral and neural dynamics.  However, when our model fails to accurately capture the stimulus dynamics that are truly driving behavioral and neural responses, our approach necessarily yields an incomplete characterization of the neural basis of the processes we are studying.

@@

Our approach also draws inspiration from the growing field of word embedding models.  The topic models~\citep{BleiEtal03} we used to embed text from the episode annotations and participants' recall transcripts are just one of many models that have been studied in an extensive literature.  The earliest approaches to word embedding, including latent
semantic analysis~\citep{LandDuma97}, used word co-occurrence statistics (i.e., how often pairs of words occur in the same documents contained in the corpus) to derive a unique feature vector for each word.  The feature vectors are constructed so that words that co-occur more frequently have feature vectors that are closer (in Euclidean distance).  Topic models are essentially an extension of those early models, in that they attempt to explicitly model the underlying causes of word co-occurrences by automatically identifying the set of themes or topics reflected across the documents in the corpus.  More recent work on these types of semantic models, including word2vec~\citep{MikoEtal13}, the Universal Sentence Encoder~\citep{CerEtal18}, GPT-2~\citep{RadfEtal19}, and GTP-3~\citep{BrowEtal20} use deep neural networks to attempt to identify the deeper conceptual representations underlying each word.  Despite the growing popularity of these sophisticated deep learning-based embedding models, we chose to prioritize interpretability of the embedding dimensions (e.g., Fig.~\ref{fig:topics}) over raw performance (e.g., with respect to some predefined benchmark).  Nevertheless, we note that our general framework is, in principle, robust to the specific choice of language model as well as other aspects of our computational pipeline.  For example, the word embedding model, timeseries segmentation model, and the episode-recall matching function could each be customized to suit a particular question space or application.  Indeed, for some questions, interpretability of the embeddings may not be a priority, and thus other text embedding approaches (including the deep learning-based models described above) may be preferable.  Further work will be needed to explore the influence of particular models on our framework's predictions and performance.

>>>>

\begin{document}
\begin{titlepage}
  \maketitle
  \thispagestyle{empty}
  \end{titlepage}

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  One potential explanation for this finding is that the topic models, trained only on episode annotations, do not capture topic proportions in participants' ``held-out'' recalls as accurately.  A second possibility is that, whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/searchlights}
\caption{\small \textbf{Brain structures that underlie the transformation of experience into memory.} \textbf{A.} We isolated the proximal diagonals from the upper triangle of the episode correlation matrix, and applied this same diagonal mask to the voxel response correlation matrix for each cube of voxels in the brain. We then searched for brain regions whose activation timeseries consistently exhibited a similar proximal correlational structure to the episode model, across participants.  \textbf{B.} We used dynamic time warping \citep{BernClif94} to align each participant's recall timeseries to the TR timeseries of the episode.  We then computed the temporal correlation matrix of each participant's warped recalls.  Next, we applied the same diagonal mask used in Panel A to isolate the proximal temporal correlations and searched for brain regions whose activation timeseries for an individual consistently exhibited a similar proximal correlational structure to each individual's recalls.  \textbf{C.} We identified a network of regions sensitive to the narrative structure of participants' ongoing experience.  The map shown is thresholded at $p < 0.05$, corrected.  The top ten \texttt{Neurosynth} terms displayed in the panel were computed using the unthresholded map.  \textbf{D}. We also identified a network of regions sensitive to how individuals would later structure the episode's content in their recalls.  The map shown is thresholded at $p < 0.05$, corrected.  The top ten \texttt{Neurosynth} terms displayed in the panel were computed using the unthresholded map.}
\label{fig:brainz}
\end{figure}

@@

The neural analyses we carried out (Fig.~\ref{fig:brainz}) also leveraged our geometric framework for characterizing the shapes of the episode and participants' recountings.  We identified one network of regions whose responses tracked with temporal correlations in the conceptual content of the episode (as quantified by topic models applied to a set of annotations about the episode).  This network included orbitofrontal cortex, ventromedial prefrontal cortex, and striatum, among others.  As reviewed by \cite{RangRitc12}, several of these regions are members of the \textit{anterior temporal system}, which has been implicated in assessing and processing the familiarity of ongoing experiences, emotions, social cognition, and reward.  A second network we identified tracked with temporal correlations in the idiosyncratic conceptual content of participants' subsequent recountings of the episode.  This network included occipital cortex, extrastriate cortex, fusiform gyrus, and the precuneus.  Several of these regions are members of the \textit{posterior medial system}~\citep{RangRitc12}, which has been implicated in matching incoming cues about the current situation to internally maintained \textit{situation models} that specify the parameters and expectations inherent to the current situation~\citep[also see][]{ZackEtal07, ZwaaRadv98}.  Taken together, our results support the notion that these two (partially overlapping) networks work in coordination to make sense of our ongoing experiences, distort them in a way that links them with our prior knowledge and experiences, and encodes those distorted representations into memory for our later use.  Our work also provides a potential framework for modeling and elucidating \textit{memory schemas}-- i.e., cognitive abstractions that may be applied to multiple related experiences~\citep[e.g.,][]{GilbMarl17, BaldEtal18}.  For example, the event-level geometric scaffolding of an experience (e.g., Fig.~\ref{fig:trajectory}A) might reflect its underlying schema, and experiences that share similar schemas might have similar shapes.  This could also help explain how brain structures including the ventromedial prefrontal cortex~\citep[Fig.~\ref{fig:brainz}; also see][]{GilbMarl17} might acquire or apply schema knowledge across different experiences (i.e., by learning patterns in the schema's shape).

Our general approach draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  Our approach to identifying neural responses to naturalistic stimuli (including experiences) entails building an explicit model of the stimulus dynamics and searching for brain regions whose responses are consistent with the model~\citep[also see][]{HuthEtal12, HuthEtal16}.  Building an explicit model of these dynamics also enables us to match up different people's recountings of a common shared experience, despite individual differences~\cite[also see][]{GagnEtal20}.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have presented a clever alternative approach: rather than building an explicit stimulus model, these studies instead search for brain responses to the stimulus that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time~\citep[also see][]{SimoChan20}.  These purely brain-driven approaches are well suited to identifying which brain structures exhibit similar stimulus-driven responses across individuals.  Further, because neural response dynamics are observed data (rather than model approximations), such approaches do not require a detailed understanding of which stimulus properties or features might be driving the observed responses.  However, this also means that the specific stimulus features driving those responses are typically opaque to the researcher.  Our approach is complementary.  By explicitly modeling the stimulus dynamics, we are able to relate specific stimulus features to behavioral and neural dynamics.  However, when our model fails to accurately capture the stimulus dynamics that are truly driving behavioral and neural responses, our approach necessarily yields an incomplete characterization of the neural basis of the processes we are studying.

@@

Our approach also draws inspiration from the growing field of word embedding models.  The topic models~\citep{BleiEtal03} we used to embed text from the episode annotations and participants' recall transcripts are just one of many models that have been studied in an extensive literature.  The earliest approaches to word embedding, including latent
semantic analysis~\citep{LandDuma97}, used word co-occurrence statistics (i.e., how often pairs of words occur in the same documents contained in the corpus) to derive a unique feature vector for each word.  The feature vectors are constructed so that words that co-occur more frequently have feature vectors that are closer (in Euclidean distance).  Topic models are essentially an extension of those early models, in that they attempt to explicitly model the underlying causes of word co-occurrences by automatically identifying the set of themes or topics reflected across the documents in the corpus.  More recent work on these types of semantic models, including word2vec~\citep{MikoEtal13a}, the Universal Sentence Encoder~\citep{CerEtal18}, GPT-2~\citep{RadfEtal19}, and GTP-3~\citep{BrowEtal20} use deep neural networks to attempt to identify the deeper conceptual representations underlying each word.  Despite the growing popularity of these sophisticated deep learning-based embedding models, we chose to prioritize interpretability of the embedding dimensions (e.g., Fig.~\ref{fig:topics}) over raw performance (e.g., with respect to some predefined benchmark).  Nevertheless, we note that our general framework is, in principle, robust to the specific choice of language model as well as other aspects of our computational pipeline.  For example, the word embedding model, timeseries segmentation model, and the episode-recall matching function could each be customized to suit a particular question space or application.  Indeed, for some questions, interpretability of the embeddings may not be a priority, and thus other text embedding approaches (including the deep learning-based models described above) may be preferable.  Further work will be needed to explore the influence of particular models on our framework's predictions and performance.
